---
title: "K-Nearest Neighbors (KNN)"
format: html
editor: visual
---

## Introduction

In this module, we will explore how the **K-Nearest Neighbors (KNN)** algorithm works and apply it to classify fruits using physical characteristics like **mass**, **width**, **height**, and **color score**.

K-Nearest Neighbors (KNN) is a simple, yet effective, algorithm used for both **classification** and **regression** tasks.

------------------------------------------------------------------------

## How KNN Works

### 📌 Core Steps

1.  **Choose** the number of neighbors `K` (usually an odd number like 3, 5, or 7).

    The first thing you do is pick how many neighbors you want the algorithm to consider when it makes a prediction. K is just a positive whole number, like 3, 5, or 7. Most people go with an odd number to avoid ties when you’re voting on the most common class, so you don’t end up with a tie as to what your result is. A popular trick is to use the square root of the total number of data points in your dataset to pick K. A smaller K means the model is more sensitive to little changes and might get thrown off by outliers. A bigger K value smooths things out a bit more and makes the model more stable, but if K is too large, it can start blending things together that maybe shouldn’t be. So picking the right K really depends on the data.

2.  **Calculate** distance (e.g., Euclidean) from the new point to all existing data points.

    Once you’ve chosen K, the next step is to figure out how close your new, unknown data point is to all the other points in your training data. This is where distance metrics come in. The most common one is Euclidean distance, which is just the straight-line distance between two points. You could also use Manhattan distance, which moves along a grid, or Minkowski distance, which is a more general formula. Essentially, you’re measuring how similar or different the new point is from everything else based on the features you’re tracking, like weight, height, or whatever.

3.  **Identify** the `K` closest neighbors.

    Once all the distances are calculated, you sort the training data by how close each point is to your new one. Then you just take the K points that are closest. These are the neighbors that are going to help decide what label or value the new point gets. The idea is that points that are close together in this space are likely to be similar.

4.  **Predict** by majority vote (for classification) or average value (for regression).

    Now that you have your K neighbors, you use them to actually make a prediction. If you’re doing classification, you look at the labels of the K neighbors and go with the one that shows up the most. Like if you’re trying to classify a mystery fruit and 3 of the 5 nearest neighbors are apples, then you’ll probably call it an apple. If you’re doing regression, which is when you're trying to predict a number, you just take the average of the neighbors’ values and use that as your prediction for the new point.

> **Example:**\
> If you're given data on the size and color of known apples, oranges, lemons, etc., and then introduced to a new "mystery fruit", KNN can tell you what kind of fruit it most likely is — based on how similar it is to others.

------------------------------------------------------------------------

## Euclidean Distance

KNN relies heavily on distance. The most commonly used metric is Euclidean distance for continuous variables, but you could also use others like Manhattan distance.

For two points A = (x₁, y₁) and B = (x₂, y₂), $d = \sqrt{(x_2 - x_1)^2 + (y_2 - y_1)^2}$, the **Euclidean distance** is: $d(A,B) = \sqrt{(x_1 - x_2)^2 + (y_1 - y_2)^2 + ... + (n_1 - n_2)^2}$

In multi-feature datasets, each feature is a dimension in the space.

------------------------------------------------------------------------

## KNN in Action: Example Codes

Install package once (if needed) `install.packages("class")`

```         
library(class)
```

#### Training data (2D)

```         
X_train <- matrix(c(1, 2, 2, 3, 3, 1, 6, 5, 7, 7), ncol = 2, byrow = TRUE)

y_train <- factor(c("A", "A", "A", "B", "B"))
```

#### New point to classify

```         
X_test <- matrix(c(3, 2), ncol = 2)
```

#### Run KNN (k = 3)

```         
predicted_label <- knn(train = X_train, test = X_test, cl = y_train, k = 3)

paste("Predicted label:", predicted_label) # output the prediction
```

------------------------------------------------------------------------

## Choosing K, kNN Cross Validation

The choice of K significantly affects the algorithm's performance. A small K can lead to overfitting (sensitive to noise in the data), while a large K can smooth out the decision boundary, possibly underfitting. Therefore, kNN cross-validation is often used to find the optimal K. Now, let's try multiple K values and use cross-validation.

#### Example codes

```         
# Load necessary package
install.packages("class")  # run this once if not already installed
library(class)

# Training data (2D points)
X_train <- matrix(c(1, 2,
                    2, 3,
                    3, 1,
                    6, 5,
                    7, 7), 
                  ncol = 2, byrow = TRUE)

# Labels for training data
y_train <- factor(c("A", "A", "A", "B", "B"))

# New data point to classify
X_test <- matrix(c(3, 2), ncol = 2)

# Run KNN (k = 3)
predicted_label <- knn(train = X_train, test = X_test, cl = y_train, k = 3)

# Output the prediction
print(paste("Predicted label:", predicted_label))
```

## 🥝 Real Dataset: Fruit Sorting with KNN

#### Dataset Loading and Setup

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(class)
library(caret)

fruit_data <- read.delim("fruit_data_with_colors.txt", sep="\t", header=TRUE)

fruit_data$fruit_label <- as.factor(fruit_data$fruit_label)
fruit_data$fruit_name <- as.factor(fruit_data$fruit_name)
```

#### Data Exploration

```{r}
summary(fruit_data)
table(fruit_data$fruit_name)

ggplot(fruit_data, aes(x = width, y = height, color = fruit_name)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Fruit Types by Size") +
  theme_minimal()
```

#### Feature Normalization and Splitting

```{r}
features <- fruit_data[, c("mass", "width", "height", "color_score")]

preproc <- preProcess(features, method = c("center", "scale"))
features_scaled <- predict(preproc, features)

set.seed(123)
train_idx <- createDataPartition(fruit_data$fruit_label, p = 0.7, list = FALSE)
train_x <- features_scaled[train_idx, ]
test_x <- features_scaled[-train_idx, ]
train_y <- fruit_data$fruit_label[train_idx]
test_y <- fruit_data$fruit_label[-train_idx]
```

#### Model Evaluation: Choosing Best K

```{r}
k_values <- seq(1, 15, 2)
accuracy_list <- sapply(k_values, function(k) {
  pred <- knn(train_x, test_x, train_y, k)
  mean(pred == test_y)
})

plot(k_values, accuracy_list, type = "b", col = "blue", pch = 19,
     xlab = "K", ylab = "Accuracy", main = "KNN Accuracy by K")
(best_k <- k_values[which.max(accuracy_list)])
```

#### Final Model and Confusion Matrix

```{r}
final_pred <- knn(train_x, test_x, train_y, best_k)

cm <- confusionMatrix(final_pred, test_y)

# Precision, Recall, F1-score by class
metrics <- data.frame(
  Class = levels(test_y),
  Precision = cm$byClass[, "Pos Pred Value"],
  Recall = cm$byClass[, "Sensitivity"],
  F1 = 2 * ((cm$byClass[, "Pos Pred Value"] * cm$byClass[, "Sensitivity"]) /
            (cm$byClass[, "Pos Pred Value"] + cm$byClass[, "Sensitivity"]))
)

metrics

# label map
label_map <- levels(fruit_data$fruit_name)
names(label_map) <- levels(fruit_data$fruit_label)

# confusion matrix transfer to heatmap
cm_df <- as.data.frame(cm$table)
cm_df$Reference <- factor(cm_df$Reference, levels = names(label_map), labels = label_map)
cm_df$Prediction <- factor(cm_df$Prediction, levels = names(label_map), labels = label_map)

# heatmap visualization
ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 6) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "KNN Confusion Matrix (Fruit Names)", x = "True Label", y = "Predicted Label") +
  theme_minimal()

```

#### 🔀 Model Comparison: KNN vs Logistic Regression

```{r}
library(nnet)  # for multinom()
library(e1071) # confusionMatrix

# using the same features in data
train_df <- data.frame(train_x)
train_df$label <- train_y
test_df <- data.frame(test_x)
test_df$label <- test_y

# Multinomial logistic regression
logit_model <- multinom(label ~ ., data = train_df)

# prediction
logit_pred <- predict(logit_model, newdata = test_df)

cm_logit <- confusionMatrix(logit_pred, test_y)

# comparison
cm_logit_df <- as.data.frame(cm_logit$table)
cm_logit_df$Reference <- factor(cm_logit_df$Reference, levels = names(label_map), labels = label_map)
cm_logit_df$Prediction <- factor(cm_logit_df$Prediction, levels = names(label_map), labels = label_map)

ggplot(cm_logit_df, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 6) +
  scale_fill_gradient(low = "white", high = "tomato") +
  labs(title = "Logistic Regression Confusion Matrix (Fruit Names)", x = "True Label", y = "Predicted Label") +
  theme_minimal()

```

#### Model Comparison Table

```{r}
comparison_df <- data.frame(
  Model = c("KNN", "Logistic Regression"),
  Accuracy = c(cm$overall["Accuracy"], cm_logit$overall["Accuracy"]),
  Kappa = c(cm$overall["Kappa"], cm_logit$overall["Kappa"])
)

comparison_df
```

------------------------------------------------------------------------

#### 🥭 Classifying a Mystery Fruit

```{r}
new_fruit <- data.frame(mass = 160, width = 7.2, height = 7.5, color_score = 0.76)
new_scaled <- predict(preproc, new_fruit)

mystery_pred <- knn(train_x, new_scaled, train_y, k = best_k, prob = TRUE)

cat("Prediction:", mystery_pred)
cat("Confidence:", attr(mystery_pred, "prob"))
```

------------------------------------------------------------------------

## KNN vs. K-Means

| Feature | KNN | K-Means |
|----|----|----|
| Supervised? | Yes | No |
| Used For | Classification / Regression | Clustering |
| Distance Use | Classify based on neighbors | Assign to closest centroid |
| Label Needed? | Yes | No |
| Type | Lazy Learner (no training phase) | Iterative optimization algorithm |

------------------------------------------------------------------------

## Strengths and Limitations

+---------------------------------------+-----------------------------------------+
| Advantages                            | Limitations                             |
+=======================================+=========================================+
| -   Easy to understand and implement  | -   Expensive at prediction time        |
|                                       |                                         |
| -   No training phase needed          | -   Sensitive to irrelevant features    |
|                                       |                                         |
| -   Flexible for multi-class problems | -   Struggles in high-dimensional space |
+---------------------------------------+-----------------------------------------+

------------------------------------------------------------------------

## Conclusion

In this module, we explored the K-Nearest Neighbors (KNN) algorithm from its theoretical foundation to practical implementation using a fruit classification task. Through both toy examples and real datasets, we gained a deeper understanding of how similarity-based classification works.

### Key Takeaways:

-   **KNN is a non-parametric, instance-based learning method** that predicts new data points by comparing their distance to known labeled data.
-   Normalization of features was critical, especially since Euclidean distance is sensitive to scale.
-   Model performance varied with the value of **K**, with `K = 1` showing perfect classification for this particular dataset. However, this could lead to overfitting in noisier or more complex data.
-   Visualization using **confusion matrices** helped us interpret model results at a glance, especially when labeled with meaningful class names (like apple, orange, etc.).
-   We compared KNN to **Logistic Regression**, a probabilistic model, and found that while KNN achieved perfect accuracy in this specific task, logistic regression showed slightly lower accuracy and Kappa. This demonstrates the potential of KNN for small, structured, and well-separated datasets.

### When to Use KNN:

KNN is particularly useful when: - You want a simple baseline model - You have a small to medium-sized dataset - Interpretability and flexibility are more important than speed

However, its limitations in terms of: - Computational cost (especially with large datasets) - Sensitivity to irrelevant features or high dimensionality should be kept in mind.

### Future Directions:

-   Try **changing the distance metric** (e.g., Manhattan, cosine) and evaluate impact.
-   Apply **dimensionality reduction** techniques like PCA to see how performance and interpretability are affected.
-   Compare KNN with other classifiers such as Decision Trees, Random Forests, or SVMs for larger datasets or more complex feature spaces.

In summary, KNN provides a solid foundation for understanding proximity-based machine learning and offers valuable intuition for feature-space dynamics in classification tasks.

------------------------------------------------------------------------

## References

-   "K-Nearest Neighbors (KNN)" - [Scikit-Learn User Guide](https://scikit-learn.org/stable/modules/neighbors.html)
-   Rahul Agarwal's [Fruit Classification on Kaggle](https://www.kaggle.com/code/rahul253801/fruits-classification/notebook)
-   Inspired by [Cluster Analysis Module](https://fuzzyatelin.github.io/bioanth-stats/module-25/module-25.html)
-   Lecture notes from BI_AN588 (2024)
